---
title: "Intro to data.table"
author: "Susanna Supalla"
date: "2024-01-16"
output: 
  ioslides_presentation:
    smaller: true
    css: presentation_style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## {data-background=tidyverse_screenshot.png data-background-size=contain}

## Another Opinion: data.table
> Data manipulation operations such as subset, group, update, join etc., are all inherently related. Keeping these related operations together allows for:
>
> -   concise and consistent syntax irrespective of the set of operations you would like to perform to achieve your end goal.
>
> -   performing analysis fluidly without the cognitive burden of having to map each operation to a particular function from a potentially huge set of functions available before performing the analysis.
>
> -   automatically optimising operations internally, and very effectively, by knowing precisely the data required for each operation, leading to very fast and memory efficient code.
>
> Briefly, if you are interested in reducing programming and compute time tremendously, then this package is for you. The philosophy that data.table adheres to makes this possible.

`rdatatable.gitlab.io/data.table/articles/datatable-intro.html`

## My favorite parts about data.table

-   Fast file reading
-   Efficient, readable syntax

## Data for Today's Examples: DIME
### Database on Ideology, Money in Politics, and Elections

- Political fundraising data
- 500 million political contributions, 1979 to 2022
- Using today: 
  - 1980 data
    - Row count: 458,213
  - 2016 data
    - Row count: 54,311,783

Bonica, Adam. 2023. Database on Ideology, Money in Politics, and Elections: Public version 3.1 [Computer file]. Stanford, CA: Stanford University Libraries. https://data.stanford.edu/dime.

## Packages for Today's Examples

-   `data.table`
-   `tidyverse`: for comparison purposes
-   `tictoc`: for capturing runtimes
-   `R.utils`: for unzipping large files


## Install Packages and Data
```{r install_load, eval = FALSE, echo = TRUE}
install.packages('data.table','tidyverse','R.utils','tictoc')
```

```{r load, echo = TRUE}
library(data.table)
library(tictoc)
```

```{r download_1980, eval = FALSE, echo = TRUE}
download.file('https://www.dropbox.com/scl/fi/gdvpzggkb0in9yruircpi/contribDB_1980.csv.gz?rlkey=rs07632m813k3g85ndek1z16g&dl=1', destfile = '~/Downloads/contribDB_1980.csv.gz')
```

Warning: this file may be larger than space available on your machine!
```{r download_2016, eval = FALSE, echo = TRUE}
#check available disk space before uncommenting this code

#download.file('https://www.dropbox.com/scl/fi/pgmoo21myed3qxsuef4rd/contribDB_2016.csv.gz?rlkey=xxc8n6lt0gck8u71w6ncyfqjp&dl=1', destfile = '~/Downloads/contribDB_2016.csv.gz')
```
## Reading Files

- `data.table`: `fread()` function
- `tidyverse`: `readr` library, `read_csv()` function
- `tidyverse`: `vroom` library, `vroom()` function
- `base R`: `read.csv()` function
  - Note: not even trying this option to avoid crashing!

## Reading Files & Timing

```{r load_1980_tbl, echo = TRUE}
tic('fread 1980')
contribdb_dt <- fread(cmd = 'gunzip -cq ~/Downloads/contribDB_1980.csv.gz')
toc(log = TRUE)

tic('read_csv 1980')
contribdb_tb <- readr::read_csv('~/Downloads/contribDB_1980.csv.gz',
                                show_col_types = F)
toc(log = TRUE)
```

## Reading Files & Timing - More

```{r load_1980_tbl2, echo = TRUE}
tic('vroom 1980')
contribdb_vr <- vroom::vroom('~/Downloads/contribDB_1980.csv.gz',
                             show_col_types = F)
toc(log = TRUE)
```

## Reading Files & Timing - Problems?

```{r load_1980_tbl3, echo = TRUE}
readr::problems(contribdb_tb)
```

## Reading Files & Timing - Looking at Problems

```{r load_1980_tbl4, echo = TRUE}
contribdb_tb[458195, 40]
contribdb_dt[458195, 40]
contribdb_vr[458195, 40]
```

## Reading Files & Timing - Repairing Problems

```{r load_1980_tbl5, echo = TRUE}
tic('vroom 1980')
contribdb_vr <- vroom::vroom('~/Downloads/contribDB_1980.csv.gz',
                             show_col_types = F,
                             guess_max = 500000)
toc(log = TRUE)
```

## Cleaning Up

```{r cleaning_up, echo = TRUE}
rm(list=ls())
gc()
```

## Reading Larger Files - fread
Warning: this file may be larger than space available on your machine!
```{r load_2016, eval = FALSE, echo = TRUE}
tic('fread 2016')
contribdb_dt <- fread(cmd = 'gunzip -cq ~/Downloads/contribDB_2016.csv.gz')
toc(log = TRUE)
```

## Cleaning Up Again

```{r cleaning_up2, echo = TRUE}
rm(list=ls())
gc()
```

## Reading Larger Files - vroom

Evaluate on your own!

```{r load_2016_2, eval = FALSE, echo = TRUE}
tic('vroom 2016')
contribdb_vr <- vroom::vroom('~/Downloads/contribDB_2016.csv.gz')
toc(log = TRUE)
```

# Working with The Data

## Loading a Subset of Data

```{r load_1980_again, eval = FALSE, echo = TRUE}

contribdb_dt <- fread(cmd = 'gunzip -cq ~/Downloads/contribDB_1980.csv.gz')
contribdb_dt
```


# Syntax

## data.table Syntax

`DT[i, j, by]`

- `i`: subset by rows
- `j`: select/compute columns
- `by`: group for computation/aggregation

## Subset Examples
```{r subset, eval = FALSE, echo = TRUE}

contribdb_dt[1500:1700,]
contribdb_dt[1500:1700] #comma not needed

contribdb_dt[c(1,6,7)]

contribdb_dt[order(amount, decreasing = TRUE)] #sort

contribdb_dt[amount > 100000] #filter
```

## Select/Compute Examples
```{r select, eval = FALSE, echo = TRUE}

contribdb_dt[, amount] #return as vector
contribdb_dt[, list(amount)] #return as data.table
contribdb_dt[, c("amount")] #same syntax as data.frame

varnames_select <- c("amount") 
contribdb_dt[, ..varnames_select] #can use variable

contribdb_dt[, sum(amount)] #computation
```

## Group Aggregation Examples
```{r group, eval = FALSE, echo = TRUE}

contribdb_dt[ , sum(amount), by = list(contributor.gender)]

contribdb_dt[, sum(amount), by = .(contributor.gender)] #alternate syntax

```

# Helper Functions

## Helpers for Subsetting/Filtering
```{r subset2, eval = FALSE, echo = TRUE}

contribdb_dt[contributor.lname %like% 'trump']

contribdb_dt[date %between% c('1980-12-01','1980-12-02')]

contribdb_dt[contributor.fname %chin% c('bill','hillary')]
```

## Shortcuts for Computations
```{r select2, eval = FALSE, echo = TRUE}

contribdb_dt[, .N] #number of rows

contribdb_dt[, .SD, .SDcols = "amount"] #subset of data

contribdb_dt[, .I[amount > 100000]] #row numbers
```

## Multiple Computations with `list()` or `.()`
```{r select3, eval = FALSE, echo = TRUE}

contribdb_dt[, list(total_amount = sum(amount),
                    mean_amount = mean(amount),
                    max_amount = max(amount))]

contribdb_dt[, .(total_amount = sum(amount),
                  mean_amount = mean(amount),
                  max_amount = max(amount))]

```

## Multiple Groupings with `list()` or `.()`
```{r group2, eval = FALSE, echo = TRUE}

contribdb_dt[, list(total_amount = sum(amount),
                  mean_amount = mean(amount),
                  max_amount = max(amount)),
             by = list(contributor.gender)]

contribdb_dt[, .(total_amount = sum(amount),
                  mean_amount = mean(amount),
                  max_amount = max(amount)),
             by = .(contributor.gender)]

```

## Subsetting, Selecting, and Grouping Together
```{r subsetselectgroup, eval = FALSE, echo = TRUE}

contribdb_dt[contributor.fname %chin% c('bill','hillary') &
               date %between% c('1980-12-01','1980-12-31')
             , .(total_amount = sum(amount),
                  mean_amount = mean(amount),
                  max_amount = max(amount))
             , by = .(contributor.gender)]

contribdb_dt[contributor.fname %chin% c('bill','hillary') &
               date %between% c('1980-12-01','1980-12-31')
             , .(total_amount = sum(amount),
                  mean_amount = mean(amount),
                  max_amount = max(amount),
                 num_contributions = .N)
             , by = .(contributor.gender,
                      female = (contributor.gender == 'F'),
                      contributor.state)]
```

## Chaining Examples
```{r chain, eval = FALSE, echo = TRUE}

test <- contribdb_dt[contributor.fname %chin% c('bill','hillary') &
                       date %between% c('1980-01-01','1980-12-31')]

test
test[,.N]

contribdb_dt[contributor.fname %chin% c('bill','hillary') &
               date %between% c('1980-01-01','1980-12-31')][,.N]

contribdb_dt[contributor.fname %chin% c('bill','hillary') &
               date %between% c('1980-01-01',
                                '1980-12-31')][,.N
                                               , by = .(contributor.gender)]
```
# Comparing Syntax to `tidyverse`

## Convert a data.table to another type

- a data.frame underlies the data.table, so data.frame syntax will already work with a data.table

- a data.table can be converted to a tbl just as a data.frame can

```{r convert, eval = FALSE, echo = TRUE}

contribdb_tbl <- tibble::as_tibble(contribdb_dt)

```

## Exercise: `data.table` vs `tidyverse`

I want one data.table that summarizes the metrics below for only women named 'sarah':
  
  - a: number of contributions from sarahs (variable name `a_num_contributions`)
  
  - b: number of contributions from sarahs that are between \$50 and \$150 (`b_50to150_num_contributions`)
 
  - c: total amount of contributions from sarahs (`c_amt_contributions`)
 
  - d: total amount of contributions from sarahs that are between \$50 and \$150 (`d_50to150_amt_contributions`)
 
  - e: number of unique sarahs (`e_last_name_s_num`)
 
  - f: average number of contributions from a sarah (`f_mean_num_contributions`)
  
  - g: average total amount of a contribution from a sarah (`g_mean_amt_contributions`)

- grouped by the state of the 'sarah' contributors (`sarah_state`)
- ordered by `a`, then by alphabetical order of `sarah_state`


# More `data.table` Manipulation

## Adding Columns, Copying DTs

```{r new_col, eval = FALSE, echo = TRUE}

contribdb_dt[, contributor.is_sarah := contributor.fname == 'sarah']

#adding a variable and assigning to a new data.table at the same time makes the variable in the original:
contribdb_dt_cp <- contribdb_dt[, contributor.is_50to150 := amount %between% c(50,150)]
contribdb_dt_cp$contributor.is_50to150[1:5]
contribdb_dt$contributor.is_50to150[1:5]

contribdb_dt_cp2 <- contribdb_dt
contribdb_dt_cp2[, contributor.is_sarah_50to150 := contributor.is_sarah &
               contributor.is_50to150]
contribdb_dt_cp2$contributor.is_sarah_50to150[1:5]
contribdb_dt_cp$contributor.is_sarah_50to150[1:5]
contribdb_dt$contributor.is_sarah_50to150[1:5]

```

## New DTs

```{r copy2, eval = FALSE, echo = TRUE}

contribdb_dt_new_cp <- data.table(contribdb_dt)

contribdb_dt_aggregated <- contribdb_dt[, .(amount_statewide = sum(amount))
                                     , by = .(contributor.state)]

contribdb_dt_filtered <- contribdb_dt[contributor.state == '']

```

## Deleting Columns

```{r delete, eval = FALSE, echo = TRUE}

names(contribdb_dt_new_cp)

contribdb_dt_new_cp[, contributor.is_sarah_50to150 := NULL]

names(contribdb_dt_new_cp)

names(contribdb_dt)

```

# Other Blazing Fast Things

## Merging and Reshaping Tables

- `merge.data.table`
- `foverlaps`
- `join`
- `roll`
- `melt`
- `dcast`

# Further Reading

## Online Resources

- All things data.table: https://rdatatable.gitlab.io/data.table/index.html
- Installation instructions: https://github.com/Rdatatable/data.table/wiki/Installation
(Note: special instructions for Mac)

